---
title: "Impact of Severe Weather Events on Public Health and Economy"
author: "Ahmed Bestawy"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview {.tabset}

Sever weather events have huge impact on public health and economy. In this analysis we will explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to discover the impact of different events on economy and number of people lost their life and who suffered injuries.

Policies regarding these types of events depends on proper analysis to prioritize resources to face theses events

We used a dplyr work flow with help of "`stringr`, `tidyr` and `tidytext` packages. we used `ggplot2` for visualizations. NOAA is a big database with many parameters. Recorded observations are from 1950 to 2011. We concerned with public health variables (number of fatalities and injuries) and the total amount of losses with respect to different kind of events.

## Data Processing {.tabset}

### 1. Getting and loading the data

Our data is relatively large. It is compressed via a bzip2 algorithm. We used `bzfile()` with `read.csv()` function to load the data.

```{r,cache=TRUE}

data_link <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(data_link, destfile = "weather.csv.bz2")

weather_data <- read.csv(bzfile("weather.csv.bz2"))
```

### 2. Loading necessary packages

Our work flow will depend mainly on `dplyr` and `stringr` packages.

```{r, warning==FALSE,message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(tidytext)
library(lubridate)
```

### 3. Taking the subset we need for our analysis

We took the subset we need for analysis.

|              |                                              |
|--------------|----------------------------------------------|
| `EVTYPE`     | Weather event type                           |
| `FATALITIES` | Number of fatalities                         |
| `INJURIES`   | Number of injuries                           |
| `PROPDMG`    | Number multiplied by property exponent value |
| `PROPDMGEXP` | Property damage exponent                     |
| `CROPDMG`    | Number multiplied by crops exponent value    |
| `CROPDMGEXP` | Crop damage exponent                         |
| `BGN_DATE`   | Date of beginning of an event                |

: Variables

We used regular expressions with `str_detect()`,`str_to_lower()` and `str_trim()` functions to combine weather events into their main categories. The original data set encoded money losses into characters (`k` for thousands, `m` for millions, etc..). We replaced these exponent characters with their actual values. Some exponents are encoded by symbols ( `=`. `-`, `?`, `+`).We treated it as a multiplication of one so we don't loss the actual value recorded for an event.

```{r, cache=TRUE}
data_subset <- weather_data %>% select(EVTYPE,BGN_DATE,FATALITIES,INJURIES, 
                                       PROPDMG,PROPDMGEXP,
                                       CROPDMG,CROPDMGEXP) %>%
    mutate(event_type = str_to_lower(EVTYPE),
           year = as.character(year(mdy_hms(BGN_DATE)))) %>%
    mutate(event_type = case_when(
        str_detect(event_type, "thunder|tstm|microburst|tunder|thuder|thuner|thundeerstorm winds|tropical|gustnado|downburst") ~ "thunder/tropical storm",
        str_detect(event_type, "tornado|torndao|landspout") ~ "tornado",
        str_detect(event_type, "heat") ~ "heat",
        str_detect(event_type, "flood|flooding|floods|floodin|urban/sml stream fld|ice jam") ~ "flood",
        str_detect(event_type, "lightning|lighting|ligntning") ~ "lightning",
        str_detect(event_type, "ice storm|glaze|ice/strong winds|freezing rain|ice roads|ice floes|ice") ~ "ice storm",
        str_detect(event_type, "frost|early frost|frost\freeze|freeze") ~ "frost/freeze",
        str_detect(event_type, "blizzard/winter storm|high wind/blizzard") ~ "blizzard",
        str_detect(event_type, "winter storms|winter storm high winds|winter weather") ~ "winter storm",
        str_detect(event_type, "wind|winds") ~ "high wind",
        str_detect(event_type, "small hail") ~ "hail",
        str_detect(event_type, "hurricane") ~ "hurricane",
        str_detect(event_type, "snow") ~ "heavy snow",
        str_detect(event_type, "fog") ~ "fog",
        str_detect(event_type, "rip currents") ~ "rip current",
        str_detect(event_type, "tropical storm") ~ "tropical storm",
        str_detect(event_type, "heavy rain|heavy shower|heavy precipitation") ~ "heavy rain",
        str_detect(event_type, "storm surge/tide|coastal surge") ~ "storm surge",
        str_detect(event_type, "hail") ~ "hail",
        str_detect(event_type, "rain|gusty wind/hvy rain|hvy rain|hvy rainrecord rainfall|rainstorm") ~ "heavy rain",
        str_detect(event_type, "cold") ~ "extreme cold",
        str_detect(event_type, "landslides") ~ "landslide",
        str_detect(event_type, "strong winds") ~ "strong wind",
        str_detect(event_type, "excessive wetness") ~ "drought", 
        str_detect(event_type, "extreme windchill|extreme wind chill|cool and wet") ~ "extreme cold",
        str_detect(event_type, "fire") ~ "wildfire",
        str_detect(event_type, "swells|high seas|surf") ~ "high surf",
        str_detect(event_type, "erosion|coastal storm|high tides") ~ "coastal flood",
        str_detect(event_type, "urban|dam break|high water") ~ "flash flood",
        str_detect(event_type, "apache county|marine accident|severe turbulence|unseasonably warm") ~ "other",
        str_detect(event_type, "wintry|mix|icy roads") ~ "winter weather",
        TRUE ~ event_type),
        property_exponent = str_trim(str_to_lower(PROPDMGEXP)),
        crop_exponent =  str_trim(str_to_lower(CROPDMGEXP))) %>%
    mutate(property_exponent = case_when(
        property_exponent ==  "k" ~  10^3,
        property_exponent == "m" ~  10^6,
        property_exponent == "b" ~  10^9,
        property_exponent == "h" ~  10^2,
        property_exponent == "8" ~  10^8,
        property_exponent =="7" ~  10^7,
        property_exponent =="6" ~  10^6,
        property_exponent =="5" ~  10^5,
        property_exponent =="4" ~  10^4,
        property_exponent =="3" ~  10^3,
        property_exponent == "2" ~  10^2,
        property_exponent == "1" ~  10^1,
        property_exponent == "0" ~  10^0,
        property_exponent == "-" ~  1,
        property_exponent == "?" ~  1,
        property_exponent == "+" ~  1,
        property_exponent == ""  ~  1,
        TRUE ~ 1 ),
        crop_exponent = case_when(
            crop_exponent == "k" ~  10^3,
            crop_exponent == "m" ~  10^6,
            crop_exponent == "b" ~  10^9,
            crop_exponent == "h" ~  10^2,
            crop_exponent == "8" ~  10^8,
            crop_exponent == "7" ~  10^7,
            crop_exponent == "6" ~  10^6,
            crop_exponent == "5" ~  10^5,
            crop_exponent == "4" ~  10^4,
            crop_exponent == "3" ~  10^3,
            crop_exponent == "2" ~  10^2,
            crop_exponent == "1" ~  10^1,
            crop_exponent == "0" ~  10^0,
            crop_exponent == "-" ~  1,
            crop_exponent == "?" ~  1,
            crop_exponent == "+" ~  1,
            crop_exponent == ""  ~  1,
            TRUE ~ 1 ))

```

### 4. Health impact

We aggregated the data by weather event type then summarize it by the sum of total fatalities and injuries records per event. The data has many missing values we filtered it out of data for better analysis.

```{r}
health_impact <- data_subset %>% group_by(event_type) %>%
    summarise(num_fatal = sum(FATALITIES),
              num_injuries = sum(INJURIES),
              num_fatal_injuries = num_fatal + num_injuries) %>%
    filter(num_fatal_injuries > 0 ) %>%
    arrange(desc(num_fatal_injuries))

```

### 4.1 Plotting Public health impact by severe weather events

We transformed data first before plotting by `pivot_longer()` function from `tidyr` package to split events by impact type (an injury or a fatality).

We used `faciting` to compare between two plots side by side. `reorder_within()` and `scale_y_reordered()` from `tidytext` package are used to arrange y axis values in a descending order within the plot code for better plot interpretation.

```{r, cache=TRUE}
plot_data <- health_impact %>% slice(1:20) %>%
             pivot_longer(cols = c(num_fatal, num_injuries),
                     names_to = "impact_type",
                     values_to = "count")


health_plot <- ggplot(data = plot_data,
                      aes(x = count,
                          y = reorder_within(event_type, count, impact_type),
                          fill = impact_type)) +
               geom_col() +
               labs(title = "Public Health Impact by Severe Weather Event",
                    x = "Total Number", 
                    y = NULL, fill = "Impact type") +
               scale_fill_manual(values = c("num_injuries" = "#1b9e77",
                                          "num_fatal" = "#d95f02")) + 
               theme_minimal() +
               scale_x_continuous(labels = scales::comma) +
               facet_wrap(~ impact_type, scales = "free",
                          labeller = as_labeller(c("num_injuries" = "Injuries",
                                                   "num_fatal" = "Fatalites"))) +
               scale_y_reordered() +
               theme(plot.margin = margin(5, 10, 5, 20, "pt"),
                 legend.position = "none")
```

### 5. Economic impact

Data aggregated by event type then property and crops actual damage are calculated

```{r}
economic_impact <- data_subset %>% group_by(event_type) %>% 
    summarise(property_damage = sum(property_exponent * PROPDMG, na.rm = TRUE) ,
               crop_damage = sum(crop_exponent * CROPDMG, na.rm = TRUE),
               total_damage = property_damage + crop_damage) %>%
    arrange(desc(total_damage))
```

### 5.1 Plotting economic impact by severe weather events

Like before, we transformed it first to make 2 plots side by side with `facet_wrap()` function.

```{r, cache=TRUE}
plot2_data <- economic_impact %>% slice(1:15) %>%
    pivot_longer(cols = c(property_damage, crop_damage),
                 names_to = "impact_type",
                 values_to = "losses")


economic_plot <- ggplot(data = plot2_data,
       aes(x = losses/10^9,
           y = reorder_within(event_type, losses, impact_type),
           fill = impact_type)) +
    geom_col() +
    theme_minimal() +
    scale_x_continuous(labels = scales::comma) +
    facet_wrap(~ impact_type, scales = "free",
               labeller = as_labeller(c("crop_damage" = "Crop Damage",
                        "property_damage" = "Property Damage"))) +
    scale_y_reordered() +
    theme(plot.margin = margin(5, 10, 5, 20, "pt"),
          legend.position = "none") +
    labs(title = "Ecnomic Impact by Severe Weather Event",
         x = "Total Losses (In Billions)", 
         y = NULL, fill = "Impact type") +
    scale_fill_manual(values = c("property_damage" = "#1b9e77",
                                 "crop_damage" = "#d95f02"))


total_economic <- ggplot(data = economic_impact[1:15,],
                         aes(y = reorder(event_type, total_damage), 
                             x = total_damage/10^9)) +
                  geom_col() + 
                  labs(title = "Total losses per event",
                  x = "Total Losses (In Billions)",
                  y = NULL)

```

### 6.The most catastrophic event in each year

```{r}
top_data <- data_subset %>% group_by(year,event_type) %>%
    summarise(num_fatal = sum(FATALITIES),
              num_injuries = sum(INJURIES),
              num_fatal_injuries = num_fatal + num_injuries,
              .groups = "drop") %>%
    filter(num_fatal_injuries > 0)

top_event_year <- top_data %>% group_by(year) %>%
    slice_max(order_by = num_fatal_injuries, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(year)) %>%
    select(c(1,2))
```

### 7.The biggest damage caused by an event in a year

```{r}
economy_data_damage_year <- data_subset %>% group_by(year,event_type) %>%
    summarise(property_damage =sum(property_exponent * PROPDMG,
                                   na.rm = TRUE),
              crop_damage = sum(crop_exponent * CROPDMG, na.rm = TRUE),
              total_damage = property_damage + crop_damage,
              .groups = "drop")

top_event_damage_year <- economy_data_damage_year %>% group_by(year) %>%
    slice_max(order_by = total_damage, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(year)) %>%
    select(c(1,2))
```

## Results {.tabset}

### Plots

From 1950 to 2011, Tornado had the greatest impact on public health. It have the biggest number of fatalities and injuries from 1950 to 1993 and 2000 to 2011.

Heat was the 2nd leading cause of fatalities. It had the greatest impact in 1995, 1999, and 2006. In 2004, hurricanes had the greatest impact. In 1998 it were the floods.

Floods caused the most devastating loss from 1950 to 2011 and hurricanes was the 2nd. The biggest property damage caused by floods and the biggest crops damage by drought. Also, Floods caused a substantial damage to crops. From 1950 to 1992, Tornadoes caused the biggst damage. In general, Floods, hurricanes, and Tornadoes had a greater impact on property damage. Drought, floods, hurricanes, and ice storms had the biggest influence on crops.

```{r}
print(health_plot)

print(total_economic)

print(economic_plot)

```

### Tables

```{r}
knitr::kable(health_impact[1:10,], caption = "The biggest 10 events")

knitr::kable(top_event_damage_year, caption = "The biggest damage caused by an event in each year")

knitr::kable(top_event_year, caption = "The top event in each year ")
```
